opioid_crisis_code

### 2020-04-27 Opioid_Crisis_Code.Rmd
code w comments

---
title: "ms_5"
author: "Suruchi Ramanujan"
date: "3/20/2020"
output: html_document
---

@@

```{r}

# total deaths per year in u.s. by cancer, drug overdose, cvd

@@

@@

@@

```{r}
plot_deathsbycommondiseases <- ggplot(totaldeathsbycommondiseases, aes(x = Year, y = Deaths, color = `Cause of Death`)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_classic() +
   labs(x = "Year",
       y = "Total Number of Deaths in the United States") +
  geom_point() +
  transition_reveal(Year)

@@

```{r}

# https://wonder.cdc.gov/controller/datarequest/D77;jsessionid=3F8131C534663BDD3F79E66994FCF12E

death_by_year_and_state <- read_csv("Drug Overdose Deaths by Year and State.csv", 
                                    col_types = "lcdccddddc") %>%
  filter(`Crude Rate` != "Unreliable")

@@

@@

death_by_year_and_state$`Crude Rate` <-as.numeric(death_by_year_and_state$`Crude Rate`)

@@

```{r, message=FALSE}
#drug treatment centers

@@

# https://simplemaps.com/data/us-cities

@@

@@

# not in final

ggplot(data = treatment_locations_map,
            mapping = aes(x = long, y = lat.y,
                          fill = count, group = group)) + 
  geom_polygon(color = "gray90", size = 0.05) + 
  theme_map() +
  scale_fill_gradient(low = "white", high = "#CB454A",
                      breaks = c(0,20,40,60,80,120)) +
    guides(fill = guide_legend(nrow = 1)) + 
    theme(legend.position = "bottom") +
  labs(title = "Distribution of Opioid Treatment Centers in 2020",
       color = "Number of Treatment Centers in State")

@@

```{r}
# in final

@@

```{r, message = FALSE}
#MA specific data

madeathbyagegender <- read_csv("MAFatalOpioidOverdosesByAgeAndGender.csv",
                               col_types = "ccddd")

macountyheroin <- read_csv("MAHeroinAsPrimarySubstanceOfUse.csv",
                           col_types = "ccccc")

madeathbyrace <- read_csv("MAOpioidDeathsByRaceAndEthnicity.csv",
                          col_types = "dcdd")

matypeofsubstance <- read_csv("MAPrimarySubstanceofUseWhenEnteringTreatmentbyTown.csv",
                              col_types = "cccccccccccc")

toxdeath <- read_csv("MAProportionDecedentsWithRxHistorybyToxicologyScreen.csv",
         skip = 1,
         col_types = "cdccccccdc")

```{r}
mavsus_death <- read_csv("MAAgeAdjustedOpioidRelatedDeathRateByYear.csv", col_types = "cdc")

mavsus_death <- mavsus_death %>%
  group_by(Geography) %>%
  filter (Year != "2015") %>%
  filter (Year != "1999") %>%
  rename(deathperhundredthousand =`Age-Adjusted Opioid-Related Death Rate per 100,000 People`)

@@

```{r}
counties <- read_csv("zip_codes_states.csv",
                     col_types = "dddccc") %>%
  filter(state == "MA") %>%
  rename(County = county) %>%
  rename(Municipality = city)

madeathbycounty <- read_csv("MAAverageAnnualOpioidRelatedDeathRateper100,000People.csv",
                            col_types = "cdcdddc")

@@

@@

@@

@@

# https://www.indexmundi.com/facts/united-states/quick-facts/massachusetts/percent-of-people-of-all-ages-in-poverty#table

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

# don't put these in final app

@@

```{r}

#treatment centers in MA

@@

@@

# need to figure this out bc color isnt working
 # addLegend(position = "bottomright", colors = "#0000FF", title = "Opioid Treatment Sites")

@@

```{r}

# models generated from data from kff.org

# ageopioid

@@

@@

@@

@@

@@

```{r}
# filtering data by race to show effects of time on deaths in each race group

@@

death_vs_white %>%
    lm(opioid_deaths ~ year, data = .) %>% 
    tidy(conf.int = TRUE) %>% 
    select(term, estimate, conf.low, conf.high) %>%
    rename(Coefficient = estimate) %>%
    rename(`Lower End` = conf.low) %>%
    rename(`Upper End` = conf.high) %>%
    gt()

@@

```{r}

@@

# showing effects of being in each age group on number of deaths idt this makes sense

>>>>

---
title: "opioid_crisis_code"
author: "Suruchi Ramanujan"
date: "4/27/2020"
output: html_document
---

@@

```{r commondiseases, echo = FALSE}

# total deaths per year in u.s. by cancer, drug overdose, cvd compared to 1999

@@

# this is to look at the total levels of deaths based on disease; pivotted
# longer for ggplot

@@

# just the info needed to map compared to 1999 proportions

@@

```{r ggplot commondiseases, echo = FALSE}

# plot year vs deaths for different diseases. transitions no longer work w
# plotly

plot_deathsbycommondiseases <- ggplot(totaldeathsbycommondiseases, aes(x = Year, 
                                        y = Deaths, color = `Cause of Death`)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_classic() +
   labs(x = "Year",
       y = "Total Number of Deaths in the United States") +
  geom_point() +
  transition_reveal(Year)

# plot year vs proportion of 1999 deaths for different diseases. transitions no
# longer work w plotly

@@

```{r overdose_counts, echo = FALSE}

# data from wonder from cdc..got rid of unreliable data

death_by_year_and_state <- read_csv("Drug Overdose Deaths by Year and State.csv"
, col_types = "lcdccddddc") %>%
  filter(`Crude Rate` != "Unreliable")

# state info for map

@@

# only want death by drugs not alcohol or unknown. join w state info

@@

# change cruderate for plot

death_by_year_and_state$`Crude Rate` <-
  as.numeric(death_by_year_and_state$`Crude Rate`)

# make a map of opioid death rates by state

@@

```{r notinuse, echo = FALSE}

#drug treatment centers

@@

# https://simplemaps.com/data/us-cities longlatinfo

@@

# treatment location info

@@

# modify data to be per capita

@@

```{r treatmentcenters, echo = FALSE}

# not in final but map of distribution of treatment centers per capita

@@

```{r madeathrates, echo = FALSE}

# read in opioid related death rate per year of ma vs that of us

mavsus_death <- read_csv("MAAgeAdjustedOpioidRelatedDeathRateByYear.csv", 
                         col_types = "cdc")

# only looking at data for which there is complete info

mavsus_death <- mavsus_death %>%
  group_by(Geography) %>%
  filter (Year != "2015") %>%
  filter (Year != "1999") %>%
  rename(deathperhundredthousand =
           `Age-Adjusted Opioid-Related Death Rate per 100,000 People`)

# ggplot grouped by geography

@@

```{r plot_1_ma, echo = FALSE}

# county data

counties <- read_csv("zip_codes_states.csv",
                     col_types = "dddccc") %>%
  filter(state == "MA") %>%
  rename(County = county) %>%
  rename(Municipality = city)

# opioid related deaths by county

madeathbycounty <- read_csv(
  "MAAverageAnnualOpioidRelatedDeathRateper100,000People.csv",
                            col_types = "cdcdddc")

# county population for per capita info

@@

# counties + longitude/latitude

@@

# modifying and dividing by time and county/subregion

@@

# county map info

@@

# poverty by county (not in use as of now)

@@

# combining all mass data together

@@

# 2001-2005 poverty percent vs opioid death rate per capita by county

@@

# not in use but correlation between poverty rate and opioid death per capita

@@

# 2006-2010 poverty percent vs opioid death rate per capita by county

@@

# 2011-2015 poverty percent vs opioid death rate per capita by county

@@

# poverty percent by county (not in use)

@@

# ggplot of opioid death rate per capita by county between 01 and 05

@@

# ggplot of opioid death rate per capita by county between 06 and 10

@@

# ggplot of opioid death rate per capita by county between 11 and 15

@@

# don't put these in final app bc they are the same graphs as above except w
# total death info rather than per capita- will prob confuse reader

@@

```{r leaflet, echo = FALSE}

# just looking at treatment centers in MA

@@

#treatment centers in MA distribution using leaflet

@@

@@

```{r ageandrace, echo = FALSE}

# models generated from data from kff.org which took data from cdc

# age vs opioid deaths over time

@@

# race vs opioid deaths over time

@@

# combine the two for ggplot later

@@

# ggplot for age over time vs opioid deaths

@@

# ggplot for race over time vs opioid deaths

@@

```{r race_lm, echo = FALSE}

# filtering data by race to show effects of time on deaths in each race group

@@

# all simple linear regression model

death_vs_white %>%
    lm(opioid_deaths ~ year, data = .) %>% 
    tidy(conf.int = TRUE) %>% 
    select(term, estimate, conf.low, conf.high) %>%
    rename(Coefficient = estimate) %>%
    rename(`Lower End` = conf.low) %>%
    rename(`Upper End` = conf.high)

@@

```{r age_lm, echo = FALSE}

@@

# showing effects of being in each age group on number of deaths idt this makes
# sense so not including it yet
